Results_setup

### 2020-04-26 doc/MatMeth.Rmd
more edits to material methods

#set wd
#load datafile
load(file="~./MLH1repo/data/Results_setup_4.21.20.RData")

@@

We tested for **effects** on mean MLH1 counts per cell; across mulitple mouse rooms, ages of mature males and differnces between embryos and neonates. No effects were found (Supplemental section).

@@

Mean DMC1 foci per cell for a single juvenile mouse from a subset of strains; _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ was quantified. Differences in DMC1 counts per cell were tested by ttests. (lateral elements of) Synaptonemal complex morphology and CREST/centromere number was used to stage spermatocytes. Early zygotene cells were defined as having a larger degree of indiviual centromeres and mostly complete axis signal. Late zygotene cells were identified as more paired centromeres and homolog axis (but below full synapsis).

@@

Mean **MLH1** foci counts were quantified from cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct, clear MLH1 foci. A quality score, between 1 to 5, was assigned to each cell based on staining quality and general spread of bivalents. Scores of 1 and 2 have the highest quality while those coded as a score of 5 had the lowest quality and were excluded from the final analysis. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure). MLH1 foci located on the XY in spermatocytes were excluded from this study. **(mouse quality) --- average of X cells quantified per female and male mice respectively - only included mice with at least 12? 10 cells quantified.**

To characterize the SC length and chromatin compaction, two image analysis algorithms were used to quantify SC area for individual cells [@wang_2019_sc] and for individual bivalents [@peterson2019]. Both apply a 'skeletonizing' transformation to the synapsed chromosomes which produces a single pixel wide 'trace' of the original chromosome shape. Hand measures of pachtyene chromosomes were performed by using ImageJ/Fiji (v1.52) [@fiji]. To remove erroneous SC isolation from the total SC algorithm, outliers were visually assessed at the mouse level and removed from the data set. Mouse means were calculated from cell-wide total SC lengths in 2984 --  `r  formatC( length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark="," )` out of 3680 -- `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name),format="d", big.mark=",")` cells with MLH1 counts (Figure X).

The DNA Cross Over algorithm isolates single straightened bivalents from an image, and returns the SC length, and locations of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents.

In this data set, isolation rates per cell (the number of all bivalents per cell) range from `r round( min(algorithm.stat.category$mean_pass.rate), 4 )` (molossinus^MSM^ male) to `r round( max(algorithm.stat.category$mean_pass.rate), 4)` (_musculus^KAZ^_ female). A curation step was applied to the total single bivalent data set to remove poor measures from the algorithm.

From the total set of cell images, `r  formatC(length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ), format="d", big.mark=",")` -- 10458 bivalent objects were isolated by the image analysis software. After a human curation step , `r formatC( length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` --- 9829 single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. From the curated single bivalent data set the proportions of bivalents by crossover number were quantified and tested by chi-square tests with the prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures.

The 'short bivalent' data set included the four or five shortest bivalents, and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 - `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A 'long bivalent' data set was calculated as above the 4th quartille in SC lengths per cell. A total of `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="male") )$n_obs)` and `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="female") )$n_obs)` were isolated from `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="male") )$ncells)` spermatoctyes and `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes respectively.

-MLH1 traits, SC based traits, CO position along single bivalents

@@

**slightly expand on the other traits**

For indiviual mice the MLH1 mean, variance and coefficicent of variance per cell, five other related meiotic traits were investigated.

We quantified mouse averages of SC based features including the 
mean total SC area per cell and the mean SC lengths for single bivalents.

mouse average of inter-focal distances IFD^norm^ and IFD^raw^ for double crossover chromosomes, and average position of crossover for single crossover bivalents.

These traits were choosen because they are markers of processes in the recombinaiton pathway and provide more detailed information about the recombination landscape. In order to examine crossover interference, inter-focal distance (IFD) was measured from the single bivalents containing two crossovers. Normalized measures were calculated by the position divided by the length of the SC for that single bivalent.

>>>>

#set wd
#load datafile
load(file="~./MLH1repo/data/Results_setup_4.24.20.RData")
#MLH1_data_setup_4.26.20.RData

@@

<We tested for **effects** on mean MLH1 counts per cell; across mulitple mouse rooms, ages of mature males and differnces between embryos and neonates. No effects were found (Supplemental section).>

@@

Mean DMC1 foci per cell for a single juvenile mouse from a subset of strains; _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ was quantified. Differences in DMC1 counts per cell were tested by ttests. The lateral elements of the synaptonemal complex morphology and CREST/centromere number was used to stage spermatocytes. Early zygotene cells were defined as having a larger degree of indiviual centromeres and mostly complete axis signal. Late zygotene cells were identified as more paired centromeres and homolog axis (but below full synapsis).

@@

Mean **MLH1** foci counts were quantified from cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct, clear MLH1 foci. A quality score, between 1 to 5, was assigned to each cell based on staining quality and general spread of bivalents. Scores of 1 and 2 have the highest quality while those coded as a score of 5 had the lowest quality and were excluded from the final analysis. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure). MLH1 foci located on the XY in spermatocytes were excluded from this study. An average of `r round( mean( ( AP_mouse_table.HQ %>% filter(sex == "female") )$Ncells), 4)` and `r round( mean( ( AP_mouse_table.HQ %>% filter(sex == "male") )$Ncells), 4)` cells were quantified for female and male mice respectively.

To characterize the SC length and chromatin compaction, two image analysis algorithms were used to quantify SC area for individual cells [@wang_2019_sc] and for individual bivalents [@peterson2019]. Both apply a 'skeletonizing' transformation to the synapsed chromosomes which produces a single pixel wide 'trace' of the original chromosome shape. Hand measures of pachtyene chromosomes were performed by using ImageJ/Fiji (v1.52) [@fiji]. To remove erroneous SC isolation from the total SC algorithm, outliers were visually assessed at the mouse level and removed from the data set. Mouse means were calculated from cell-wide total SC lengths in  `r  formatC( length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark="," )` out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name),format="d", big.mark=",")` cells with MLH1 counts (Figure X).

The DNA Cross Over algorithm isolates single straightened bivalents from an image, and returns the SC length, and locations of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this data set, isolation rates per cell (the number of all bivalents per cell) range from `r round( min(algorithm.stat.category$mean_pass.rate), 3)` (molossinus^MSM^ male) to `r round( max(algorithm.stat.category$mean_pass.rate), 3)` (_musculus^KAZ^_ female). A curation step was applied to the total single bivalent data set to remove poor measures from the algorithm. From the total set of cell images, `r  formatC(length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ), format="d", big.mark=",")` bivalent objects were isolated by the image analysis software. After a human curation step , `r formatC( length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. From the curated single bivalent data set the proportions of bivalents by crossover number were quantified and tested by chi-square tests with the prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. The 'short bivalent' data set included the four or five shortest bivalents, and excluded the X bivalent in oocytes. A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A 'long bivalent' data set was calculated as above, using the 4th quartille for SC lengths per cell. A total of `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="male") )$n_obs)` and `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="female") )$n_obs)` bivalents were isolated from `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="male") )$ncells)` spermatoctyes and `r sum(  (Q1.real.long_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes respectively.

For indiviual mice the MLH1 mean, variance and coefficicent of variance per cell were investigated. We quantified mouse averages of SC based features including the mean total SC area per cell and the mean SC lengths for single bivalents. The crossover position traits crossover position for single crossover bivalents and crossover interference from double crossover interference. For double crossover bivalents the distance between two MLH1 foci, inter-focal distances (IFD), was examined in IFD^raw^ or the normalized by the SC length IFD^norm^. These traits were examined using a our model framework because they are markers of processes in the recombinaiton pathway and provide more detailed information about the recombination landscape.

@@

### 2020-04-26 src/Result_Setup.Rmd
more edits to material methods

load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_4.21.20.RData")
#load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_2.24.20.RData")
#load(file = "C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_1.31.20.RData")

@@

```{r Q1.rbar, echo=FALSE}

@@

@@

#calq sex mean
AST.mean.MLH1.obs <- ddply(.data=HetC.AST, 
                 .(category),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean( nMLH1.foci),
                 var.MLH1 = var( nMLH1.foci)
)
AST.HETC.val <- AST.mean.MLH1.obs$mean.MLH1[1] / AST.mean.MLH1.obs$mean.MLH1[2]

#Molosinus

#MSM
HetC.MSM <- MLH1_data[(MLH1_data$strain == "MSM")&(MLH1_data$quality < 5),]
unique(HetC.MSM$mouse)

>>>>

load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_4.26.20.RData")
#load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_2.24.20.RData")
#load(file = "C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_1.31.20.RData")

@@

```{r Q1.rbar, echo=FALSE, eval=FALSE}

@@

#no mice removed for low cell numbers?

@@

#calq sex mean
AST.mean.MLH1.obs <- ddply(.data=HetC.AST, 
                 .(category),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean( nMLH1.foci),
                 var.MLH1 = var( nMLH1.foci)
)
AST.HETC.val <- AST.mean.MLH1.obs$mean.MLH1[1] / AST.mean.MLH1.obs$mean.MLH1[2]
#Molosinus

#MSM
HetC.MSM <- MLH1_data[(MLH1_data$strain == "MSM")&(MLH1_data$quality < 5),]
#unique(HetC.MSM$mouse)
